---
title: "FIFA World Cup 2022 Analysis"
subtitle: "Final Group Project"
author: "Chenjian Wei, Riya Dsouza, Ryland Jones"
date: "May 3, 2026"
format:
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

# Load packages
library(tidyverse)
library(knitr)
library(rvest)
library(kableExtra)
library(readxl)
library(janitor)
library(ggrepel)
library(viridis)
```

```{r}
#| label: load-data
#| message: false
#| warning: false
#| echo: false

worldCupRaw <- read_xlsx("worldCup2022.xlsx")
```

```{r}
#| label: clean-data
#| message: false
#| warning: false
#| echo: false

worldCupClean <- worldCupRaw %>%
  
  clean_names() %>%
  
  filter(
    !is.na(squad),
    squad != "Squad",
    !str_detect(squad, "Playing")
  ) %>%
  
  select(
    squad,
    players = number_pl,
    age,
    poss,
    matches_played = mp,
    starts,
    minutes = min,
    nineties = x90s,
    goals = gls_9,
    assists = ast_10,
    goals_plus_assists = g_a_11,
    goals_no_pk = g_pk_12,
    pk,
    pkatt = p_katt,
    yellow_cards = crd_y,
    red_cards = crd_r
  ) %>%
  
  mutate(across(-squad, ~as.numeric(as.character(.)))) %>%
  drop_na(age, goals_plus_assists)
```

## Introduction

This project uses player level data from the 2022 FIFA World Cup to examine how different performance variables relate to overall effectiveness. In addition to analyzing raw statistics like goals, assists, and minutes played, we will also use per-90 metrics to allow for fair comparisons between players with different playing time. The project involves cleaning and organizing the dataset, generating visualizations to compare players and trends, and computing summary statistics to identify relationships between variables.

## Tidy data

The dataset was cleaned to follow tidy data principles, where each variable is a column and each row represents a team.\
\
First, \`clean_names()\` was used to make all column names consistent and easier to use. Then, unnecessary rows were removed, including rows with missing team names and rows that were not actual teams, such as those labeled "Squad" or "Playing".\
\
Next, only important variables were kept, such as team name, age, matches played, and goals. Some columns were also renamed to make them more clear, for example, \`mp\` was renamed to \`matches_played\` and \`gls_9\` to \`goals\`.\
\
After these steps, the dataset became cleaner, easier to understand, and ready for further analysis.

## Methodology

This project uses an exploratory data analysis paradigm, since the goal is to examine World Cup data to identify patterns and trends in team performance rather than test a specific hypothesis. This approach was chosen because it allows for flexibility in analyzing the data and helps show insights through summary statistics and visualizations. For the coding work, a Tidyverse approach in R was used. This was chosen because it provides a clear and efficient workflow for data cleaning, transformation, and visualization. Tools like dplyr and ggplot2 make it easier to change the dataset and create meaningful visualizations, which supports the goals of exploratory analysis.

## Cases and Variables

In this dataset, a case represents a single World Cup team. Each row corresponds to one team and summarizes its performance statistics during the tournament.
The case attributes are the variables that describe each team’s performance. These include variables such as total goals scored, goals per 90 minutes, matches played, and other performance related statistics. These attributes are quantitative and are used to compare teams and to identify any patterns in performance.
Where applicable, the units of measurement are defined. For example, goals are counted as totals, and goals per 90 minutes adjust for playing time so teams can be compared fairly. These variables help analyze and compare team performance.


## Summary Statistics
```{r}
#| echo: false
#| label: tbl-descriptive-stats
#| tbl-cap: "Summary Statistics of Key Variables"
#| tbl-pos: H

worldCupClean %>%
  summarize(
    avg_goals = mean(goals_plus_assists, na.rm = TRUE),
    median_goals = median(goals_plus_assists, na.rm = TRUE),
    min_goals = min(goals_plus_assists, na.rm = TRUE),
    max_goals = max(goals_plus_assists, na.rm = TRUE),
    sd_goals = sd(goals_plus_assists, na.rm = TRUE),
    
    avg_age = mean(age, na.rm = TRUE),
    avg_players = mean(players, na.rm = TRUE),
    avg_poss = mean(poss, na.rm = TRUE)
  ) %>%
  kable(
    caption = "Summary Statistics of Team Performance"
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

## Plot and Table

```{r}
#| echo: false
#| label: fig-playersVsPerformance
#| fig-cap: "Top 8 team Average goals per match by team number of players"
#| fig-pos: H
#| fig-height: 5
ggplot(worldCupClean, aes(x = players, y = goals_plus_assists, color = squad)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Number of Players vs Performance",
    x = "Number of Players",
    y = "Goals + Assists",
    color = "Team"
  ) +
  geom_text_repel(
    data = worldCupClean %>% slice_max(goals_plus_assists, n = 10),
    aes(label = squad),
    size = 4,
    max.overlaps = Inf
  ) +
  scale_color_viridis_d(option = "D") +
  theme_minimal()
```

@fig-players-goals shows the relationship between the number of players and the average goals per match for the top 8 teams. From the scatter plot, there does not appear to be a strong linear relationship between team size and scoring performance. Some teams with a moderate number of players, such as England and Portugal, achieve relatively high average goals, while others with similar or even higher player counts do not perform as well. This suggests that simply having more players does not necessarily lead to better offensive outcomes.

```{r}
#| echo: false
#| label: tbl-playersVsPerformance1
#| tbl-cap: "Top 8 World Cup Teams – Average Goals Performance"
#| tbl-pos: H
avg_perf <- mean(worldCupClean$goals_plus_assists)
avg_players <- mean(worldCupClean$players)

top_table <- worldCupClean %>%
  arrange(desc(goals_plus_assists)) %>%
  mutate(
    Rank = row_number(),
    Performance_vs_Avg = round(goals_plus_assists - avg_perf, 1),
    Players_vs_Avg = round(players - avg_players, 1),
  ) %>%
  select(
    Rank,
    Team = squad,
    Players = players,
    `Goals + Assists` = goals_plus_assists,
    `Performance vs Avg` = Performance_vs_Avg,
    `Players vs Avg` = Players_vs_Avg,
  )

kable(
  top_table,
  caption = "<span style='font-size:20px; font-weight:bold; color:black;'>Offensive Performance Compared to Number of Players</span>",
  escape = FALSE
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

```{r}
#| echo: false
#| label: tbl-playersVsPerformance2
#| tbl-cap: "Top 8 World Cup Teams – Average Goals Performance"
#| tbl-pos: H
corr <- cor(worldCupClean$players, worldCupClean$goals_plus_assists)

model <- lm(goals_plus_assists ~ players, data = worldCupClean)

simple_table <- data.frame(
  Metric = c("Average Number of Players", 
             "Average Goals + Assists", 
             "Correlation (Number of Players vs Performance)", 
             "Slope (Number of Players Effect)"),
  
  Value = c(round(avg_players, 2),
            round(avg_perf, 2),
            round(corr, 2),
            round(coef(model)[2], 2))
)

kable(
  simple_table,
  caption = "Impact of The Number of Players on Offensive Performance",
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

@tbl-playersVsPerformance1 provides a clearer comparison of the top 8 teams based on their average goals per match. England ranks highest with 2.60 goals per match, followed by Portugal and France, indicating strong offensive performance. On the other hand, Morocco has the lowest average, showing a more defensive or less efficient scoring style. Overall, the table highlights that performance differences between teams are quite noticeable, and factors beyond team size—such as strategy and player quality—likely play a more important role.

```{r}
#| echo: false
#| label: fig-ageVsPerformance
#| fig-cap: "Top 8 team Average goals per match by team number of players"
#| fig-pos: H
#| fig-height: 5
ggplot(worldCupClean, aes(x = age, y = goals_plus_assists, color = squad)) +
  
  geom_point(size = 3) +
  
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  
  labs(
    title = "Age vs Performance",
    x = "Average Team Age",
    y = "Goals + Assists",
    color = "Team"
  ) +
  
  geom_text_repel(
    data = worldCupClean %>% slice_max(goals_plus_assists, n = 10),
    aes(label = squad),
    size = 4,
    max.overlaps = Inf
  ) +
  
  scale_color_viridis_d(option = "D") +
  
  theme_minimal()
```

```{r}
#| echo: false
#| label: tbl-ageVsPerformance1
#| tbl-cap: "Top 8 World Cup Teams – Average Goals Performance"
#| tbl-pos: H
avg_perf <- mean(worldCupClean$goals_plus_assists)
avg_age <- mean(worldCupClean$age)

top_table <- worldCupClean %>%
  arrange(desc(goals_plus_assists)) %>%
  mutate(
    Rank = row_number(),
    Performance_vs_Avg = round(goals_plus_assists - avg_perf, 1),
    Age_vs_Avg = round(age - avg_age, 1),
  ) %>%
  select(
    Rank,
    Team = squad,
    Age = age,
    `Goals + Assists` = goals_plus_assists,
    `Performance vs Avg` = Performance_vs_Avg,
    `Age vs Avg` = Age_vs_Avg,
  )

kable(
  top_table,
  caption = "<span style='font-size:20px; font-weight:bold; color:black;'>Top World Cup Teams- Offensive Performance and Possession Compared to Average</span>",
  escape = FALSE
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

```{r}
#| echo: false
#| label: tbl-ageVsPerformance2
#| tbl-cap: "Top 8 World Cup Teams – Average Goals Performance"
#| tbl-pos: H
corr <- cor(worldCupClean$players, worldCupClean$goals_plus_assists)

model <- lm(goals_plus_assists ~ players, data = worldCupClean)

simple_table <- data.frame(
  Metric = c("Average Number of Players", 
             "Average Goals + Assists", 
             "Correlation (Number of Players vs Performance)", 
             "Slope (Number of Players Effect)"),
  
  Value = c(round(avg_players, 2),
            round(avg_perf, 2),
            round(corr, 2),
            round(coef(model)[2], 2))
)

kable(
  simple_table,
  caption = "Impact of The Number of Players on Offensive Performance",
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

```{r}
#| echo: false
#| label: fig-possessionVsPerformance
#| fig-cap: "Top 8 team Average goals per match by team number of players"
#| fig-pos: H
#| fig-height: 5
ggplot(worldCupClean, aes(x = poss, y = goals_plus_assists, color = squad)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "Possession vs Performance",
    x = "Possession Percentage",
    y = "Goals + Assists",
    color = "Team"
  ) +
  geom_text_repel(
    data = worldCupClean %>% slice_max(goals_plus_assists, n = 10),
    aes(label = squad),
    size = 4,
    max.overlaps = Inf
  ) +
  scale_color_viridis_d(option = "D") +
  theme_minimal()
```

```{r}
#| echo: false
#| label: tbl-possessionVsPerformance1
#| tbl-cap: "Top 8 World Cup Teams – Average Goals Performance"
#| tbl-pos: H
avg_perf <- mean(worldCupClean$goals_plus_assists)
avg_poss <- mean(worldCupClean$poss)

top_table <- worldCupClean %>%
  arrange(desc(goals_plus_assists)) %>%
  mutate(
    Rank = row_number(),
    Performance_vs_Avg = round(goals_plus_assists - avg_perf, 1),
    Poss_vs_Avg = round(poss - avg_poss, 1),
  ) %>%
  select(
    Rank,
    Team = squad,
    Possession = poss,
    `Goals + Assists` = goals_plus_assists,
    `Performance vs Avg` = Performance_vs_Avg,
    `Possession vs Avg` = Poss_vs_Avg,
  )

kable(
  top_table,
  caption = "<span style='font-size:20px; font-weight:bold; color:black;'>Offensive Performance Compared to Possession Percentage</span>",
  escape = FALSE
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

```{r}
#| echo: false
#| label: tbl-possessionVsPerformance2
#| tbl-cap: "Top 8 World Cup Teams – Average Goals Performance"
#| tbl-pos: H
corr <- cor(worldCupClean$poss, worldCupClean$goals_plus_assists)

model <- lm(goals_plus_assists ~ poss, data = worldCupClean)

simple_table <- data.frame(
  Metric = c("Average Possession", 
             "Average Goals + Assists", 
             "Correlation (Possession vs Performance)", 
             "Slope (Possession Effect)"),
  
  Value = c(round(avg_poss, 2),
            round(avg_perf, 2),
            round(corr, 2),
            round(coef(model)[2], 2))
)

kable(
  simple_table,
  caption = "Impact of Possession on Offensive Performance",
) %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover", "condensed")
  )
```

## Data Source & Ethics

The data used in this project comes from publicly available World Cup statistics, which were accessed and imported into R for analysis. The dataset is clearly structured and includes variables such as teams, goals, and performance variables, making it suitable for analysis. This dataset satisfies the FAIR principles in many ways. It is findable and accessible because it is publicly available online, and it is interoperable since it can be downloaded and used in formats compatible with R. Also, it is reuseable because the data is well organized and can be used for many different types of analyses. The dataset also aligns with the CARE principles, although it is in a less direct way. Since the data represents grouped team performance rather than individual information, there are less concerns relating to privacy or ethical issues. However, one limitation is that the dataset does not provide detailed context about how the data was collected or whether there are any biases in reporting, which makes it more difficult to fully assess its ethical implications.

## Conclusion

## Reference

“2022 World Cup Stats.” *FBref.com*, Sports Reference, <https://fbref.com/en/comps/1/2022/2022-World-Cup-Stats>. Accessed 3 May 2026.

## Author Contributions

Chenjian Wei:\
Riyadsouza:\
RylandJones:

## Code Appendix

```{r}
#| label: codeAppend
#| ref-label: !expr knitr::all_labels()
#| echo: true
#| eval: false

```
