---
title: "FIFA World Cup 2022 Analysis"
subtitle: "Final Group Project"
author: "Chenjian Wei, Riyadsouza, RylandJones"
date: "May 3, 2026"
format:
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
execute:
  warning: false
  message: false
---

```{r}
#| echo: false

library(tidyverse)
library(readxl)
library(janitor)
library(ggplot2)
library(gt)

worldCupRaw <- read_xlsx("worldCup2022.xlsx")

worldCupClean <- worldCupRaw |>
  clean_names() |>
  filter(
    !is.na(squad),
    squad != "Squad",
    !str_detect(squad, "Playing")
  ) |>
  select(
    squad,
    players = number_pl,
    age,
    poss,
    matches_played = mp,
    starts,
    minutes = min,
    nineties = x90s,
    goals = gls_9,
    assists = ast_10,
    goals_plus_assists = g_a_11,
    goals_no_pk = g_pk_12,
    pk,
    pkatt = p_katt,
    yellow_cards = crd_y,
    red_cards = crd_r
  )

top8 <- worldCupClean |>
  filter(matches_played >= 5) |>
  mutate(
    goals = as.numeric(goals),
    matches_played = as.numeric(matches_played),
    age = as.numeric(age),
    avg_goals = goals / matches_played
  )

```

## Introduction

This project uses player level data from the 2022 FIFA World Cup to examine how different performance variables relate to overall effectiveness. In addition to analyzing raw statistics like goals, assists, and minutes played, we will also use per-90 metrics to allow for fair comparisons between players with different playing time. The project involves cleaning and organizing the dataset, generating visualizations to compare players and trends, and computing summary statistics to identify relationships between variables.

## Tidy data

The dataset was cleaned to follow tidy data principles, where each variable is a column and each row represents a team.\
\
First, \`clean_names()\` was used to make all column names consistent and easier to use. Then, unnecessary rows were removed, including rows with missing team names and rows that were not actual teams, such as those labeled "Squad" or "Playing".\
\
Next, only important variables were kept, such as team name, age, matches played, and goals. Some columns were also renamed to make them more clear, for example, \`mp\` was renamed to \`matches_played\` and \`gls_9\` to \`goals\`.\
\
After these steps, the dataset became cleaner, easier to understand, and ready for further analysis.

## Plot and Table

```{r}
#| echo: false
#| label: fig-age-goals
#| fig-cap: "Top 8 team Average goals per match by team average age"
#| fig-pos: H
library(tidyverse)
library(readxl)
library(janitor)
library(ggplot2)

worldCupRaw <- read_xlsx("worldCup2022.xlsx")

worldCupClean <- worldCupRaw |>
  clean_names() |>
  filter(
    !is.na(squad),
    squad != "Squad",
    !str_detect(squad, "Playing")
  ) |>
  select(
    squad,
    age,
    matches_played = mp,
    goals = gls_9
  )

top8 <- worldCupClean |>
  filter(matches_played >= 5) |>
  mutate(
    goals = as.numeric(goals),
    matches_played = as.numeric(matches_played),
    age = as.numeric(age),
    avg_goals = goals / matches_played
  )

ggplot(top8, aes(x = age, y = avg_goals, color = squad)) +
  geom_point(size = 4) +
  labs(
    x = "Average Age",
    y = "Goals per Match",
    color = "Country"
  ) +
  theme_minimal()
```

@fig-age-goals illustrates the relationship between average team age and goals per match. The scatter plot suggests that there is no clear linear relationship between age and scoring performance. While some older teams achieve high scoring rates, others do not, indicating that age alone is not a strong predictor of offensive success.

These results suggest that additional factors, such as team strategy, player skill, and match conditions, may play a more important role in determining scoring performance than average age alone.

```{r}
#| echo: false
#| label: tbl-top8
#| tbl-cap: "Top 8 World Cup Teams – Average Goals Performance"
#| tbl-pos: H

top8 |>
  mutate(Rank = row_number()) |>
  select(Rank, squad, age, avg_goals) |>
  gt() |>
  cols_label(
    squad = "Team",
    age = "Age",
    avg_goals = "Goals per Match"
  ) |>
  tab_header(
    title = "Top 8 World Cup Teams"
  ) |>
  fmt_number(
    columns = c(age, avg_goals),
    decimals = 2
  ) |>
  cols_align(align = "center") |>
  opt_row_striping()
```

@tbl-top8 summarizes the key statistics for the selected teams that played at least five matches in the 2022 FIFA World Cup. The table presents team average age and goals per match, allowing for comparison across teams with different levels of performance.

Overall, teams such as England and France show relatively high goals per match, while others like Croatia and Morocco demonstrate lower scoring output. This variation highlights differences in offensive effectiveness among the top-performing teams.

## Conclusion

## Reference

“2022 World Cup Stats.” *FBref.com*, Sports Reference, <https://fbref.com/en/comps/1/2022/2022-World-Cup-Stats>. Accessed 3 May 2026.

## Author Contributions

Chenjian Wei:\
Riyadsouza:\
RylandJones:

## Code Appendix

```{r}
#| echo: true
#| eval: false

# Style Guide: tidyverse style guide
# Primary Author: 
# Reviewer: 

# 2022 World Cup Data ----
## Get data from the 2022 World Cup and create plots showing statistics by
## team vs average team age

# Step 1: Load Packages ----
## Needed Packages: tidyverse, knitr, rvest, kableExtra, readxl
install.packages("readxl")
install.packages("gt")
library(tidyverse)
library(knitr)
library(rvest)
library(kableExtra)
library(readxl)
library(janitor)
library(dplyr)
library(gt)
library(ggplot2)
# Step 2: Load Data ----
## Needed data: World Cup Team Data

worldCupRaw <- read_xlsx("worldCup2022.xlsx")

View(worldCupRaw)
# Step 3: Tidy the data ----
worldCupClean <- worldCupRaw |>
  
  clean_names() |>
  
  filter(
    !is.na(squad),
    squad != "Squad",
    !str_detect(squad, "Playing")
  ) |>
  
  select(
    squad,
    players = number_pl,
    age,
    poss,
    matches_played = mp,
    starts,
    minutes = min,
    nineties = x90s,
    goals = gls_9,
    assists = ast_10,
    goals_plus_assists = g_a_11,
    goals_no_pk = g_pk_12,
    pk,
    pkatt = p_katt,
    yellow_cards = crd_y,
    red_cards = crd_r
  )
  
# Step 4: Make a plot ----
top8 <- worldCupClean |>
  filter(matches_played >= 5) |>
  mutate(
    goals = as.numeric(goals),
    matches_played = as.numeric(matches_played),
    age = as.numeric(age),
    avg_goals = goals / matches_played
  )

ggplot(top8, aes(x = age, y = avg_goals, color = squad)) +
  geom_point(size = 4) +
  labs(
    title = "Average Goals vs Age for Top 8 Teams",
    x = "Average Age",
    y = "Goals per Match",
    color = "Country"
  ) +
  theme_minimal()
## Step 4A: Make a table ----
top8 |>
  mutate(Rank = row_number()) |>
  select(Rank, squad, age, avg_goals) |>
  gt() |>
  cols_label(
    squad = "Team",
    age = "Age",
    avg_goals = "Goals per Match"
  ) |>
  tab_header(
    title = "Top 8 World Cup Teams - Average Goals Performance"
  ) |>
  fmt_number(
    columns = c(age, avg_goals),
    decimals = 2
  ) |>
  cols_align(align = "center") |>
  opt_row_striping()
```
